A C-to-RTL flow as an energy efficient alternative to embedded processors in digital systems

### 2010-05-24 Documents/Publications/dsd2010/paper.tex
dsd2010: fixed references for AHIR, also put all URLs in footnotes

\documentclass[conference]{IEEEtran} 
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{multirow}
\usepackage{times}

@@

\author{
  Sameer D. Sahasrabuddhe, Kunal P. Ghosh, Sreenivas
  Subramanian, Kavi Arya and Madhav P. Desai\\
  Indian Institute of Technology -- Bombay, Powai, Mumbai -- 400076, INDIA\\
  Email: \{sameerds@cse, ssreenivas@ee, kunalghosh@ee, kavi@cse, madhav@ee\}.iitb.ac.in
}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

Several commercial tools which translate a C program to hardware are
also available.  For example, Catapult-C \cite{catapult-c} from
Mentor Graphics, C-to-Silicon \cite{c-to-silicon} from Cadence,
and Synopsys HLS \cite{synopsys-hls} from Synopsys.  Currently,
we have no access to the internal details of these proprietary tools.

@@

%% Common operators such as equality $(A = B)$, concatenation $(a.b)$ and
%% prefix $(a \leq b)$ are defined to have sensible meanings. The prefix
%% operator defines a partial order on label sequences. The longest
%% common prefix (LCP) of two sequences is the longest sequence that is a
%% prefix of both the sequences.

@@

The synthesis flow uses the LLVM \cite{ref:LLVM}  framework to parse and optimise the
%
% http://www.llvm.org
%
input C program. This is translated to an AHIR specification using a
CDFG as an intermediate step.

@@

The correctness of the second step (CDFG to AHIR) is shown
as follows \cite{ahir_thesis}: suppose that
$A$ was the AHIR specification generated from the
reference CDFG $G$.  Then every execution sequence
in $A$ is a member of the set of
possible execution sequences of $G$.  Now,
from $A$, we construct a new CDFG $G'$ such
that every execution sequence of  $G'$  is
an element of the set of execution sequences
of $A$.  Now, $G'$ and $G$ are shown to be isomorphic,
so that their sets of execution sequences are in 
a one-to-one correspondence with each other.  This
shows that the sets of execution sequences in $A$ and
$G$ are in a one-to-one correspondence.

@@

In each case, the VHDL generated by
the C-to-RTL flow was synthesized using Synposys Design Compiler,
using the OSU standard cell library \cite{OSU} for the TSMC $0.18\mu$ 
%
% vcag.ecen.okstate.edu/projects/scells/
%
technology.  
Memories were modeled using CACTI SRAM models \cite{CACTI}. 
The resulting net-list was mapped to layout using
the Cadence SOC Encounter tool.  No human intervention was
used in the entire process from C to the final layout (other
than specifying the operating frequency to the automatic place-and-route
tools).   The total time needed for the entire process (from C to layout)
was of the order of a few minutes for A5/1 to a few hours for the LPK
case.
%
% Premkishore Shivakumar, Norman P. Jouppi. Cacti 3.0 : An Integreted Cache Timing, Power, and Area model, Western Research
% Laboratory, 250 University Avenue Palo Alto, California 94301 USA, August 2001
%
A simulation based toggle file was then used with the extracted
net-list to estimate power dissipation in each of the generated
circuits.
The reciprocal of the energy used for completing a task is equivalent to
the throughput achieved for each watt of power supplied, commonly
termed as {\em performance per watt} 
%
% TODO: add a comment that the critical path was through 
% adders/multipliers which were unoptimized!

@@

As a reference to judge the energy efficiency of
the AHIR generated circuits, we consider a 
popular low power processor (henceforth denoted
by $P$), namely the Intel Atom N270 processor \cite{ref:Atom},
%
% www.intel.com/products/processor/atom/index.htm
% 
which is built in Intel's $45nm$ technology.
This processor is readily available, combines high performance
with low power dissipation, has full floating point
support, and is a part of many low power
computing devices.   
The sample set of C programs was run on the processor $P$ (exactly
the same program used in the hardware generation was run on the processor).
In each case, the number of cycles needed
to finish one task (as described above) specific to that program was measured. 
The data
sets for the programs were small enough to fit into the processor
cache, and the cycle count was averaged across multiple runs.  These
cycle counts were then used to find the delay and energy consumption
for the processor.  All values for area, frequency and
power dissipation for the processor were obtained from the
corresponding data-sheet.

@@

Now, suppose
that  we scale the 
energy numbers assuming migration of the AHIR circuits
from $180nm$ to the $45nm$ technology node, using 
the following  scaling rules (assume that $S = 45/180 = 0.25$):
\begin{itemize}
\item The operating voltage is scaled from $1.8V$ to $0.9V$.
\item The internal capacitances are assumed to be scaled by $S$.
\item The operating frequency of the generated circuit is scaled
by $1/S$.
\item Half of the total power dissipation of the $45nm$ circuits
is due to leakage \cite{ref:Frank}, and the other half is due to switching power. 
Note that this is a conservative estimate, and the actual leakage is
likely to be lower at the power levels corresponding to the
AHIR circuits.
%
% 
The leakage power is a negligible fraction of the
total power dissipation in the $180nm$ technology.
\item The area is scaled by $S^2$.
\end{itemize}
Thus, the per task energy consumption due to switching
activity alone will scale by $1/16$, and since
we are assuming that leakage power dissipation is half the total,
the total per task energy consumption of the circuit will 
scale by $1/8$.
The area of the circuits will scale by $1/16$.  
The power dissipation will scale by $1/2$ (this would have been
$1/4$ without leakage).
This scaling, though approximate, can be justified since the 
area of the resulting circuits (in each case) is small, so 
that interconnect effects are not likely to be pronounced.

@@

In Table \ref{table:Ratios}, we tabulate the ratios of
the performance parameters area, frequency, delay, power
and energy for the processor $P$ relative to the scaled values
for the AHIR circuits assuming the scaling rules shown above.
From the table, we see that across all applications, the energy consumed
by AHIR circuits is considerably lower than the
processor (the energy per task for the processor is
between $4.7X$ and $116.6X$ that of the corresponding AHIR circuit).  
The improvement is highest in the case of
bit-level manipulation applications such as A5/1, and
lowest in the case of random ``pointer-chasing'' code.
The delays in completing the tasks are lower 
for the processor $P$ in almost all the cases.  This
is mainly due to the fact that currently, the operators used in 
the AHIR generated RTL are not pipelined or optimized in any
manner during the logic synthesis and physical
design. This limits the operating clock frequency of the circuits
synthesized from the AHIR generated RTL.  
%
% Need more elaboration here
%

@@

%
% Need to revisit this
%

>>>>

\documentclass[conference]{IEEEtran} 
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{multirow}
\usepackage{times}
\usepackage{url}

@@

\author{Sameer D. Sahasrabuddhe, Sreenivas
  Subramanian, Kunal P. Ghosh, Kavi Arya and Madhav P. Desai\\
  Indian Institute of Technology -- Bombay, Powai, Mumbai -- 400076, INDIA\\
  Email: \{sameerds@cse, ssreenivas@ee, kunalghosh@ee, kavi@cse,
  madhav@ee\}.iitb.ac.in }
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

Several commercial tools which translate a C program to hardware are
also available. For example,
Catapult-C\footnote{\url{http://www.mentor.com/products/esl/high_level_synthesis/}}
from Mentor Graphics,
C-to-Silicon\footnote{\url{http://www.cadence.com/products/sd/silicon_compiler/}}
from Cadence, and Synopsys
HLS\footnote{\url{http://www.synopsys.com/Tools/SLD/HLS/}} from
Synopsys. Currently, we have no access to the internal details of
these proprietary tools.

@@

@@

The synthesis flow uses the LLVM Compiler
Infrastructure\footnote{\url{http://llvm.org/}} to parse and optimise
the input C program. This is translated to an AHIR specification using
a CDFG as an intermediate step.

@@

The correctness of the second step (CDFG to AHIR) is shown
as follows\cite{ahir_thesis}: suppose that
$A$ was the AHIR specification generated from the
reference CDFG $G$.  Then every execution sequence
in $A$ is a member of the set of
possible execution sequences of $G$.  Now,
from $A$, we construct a new CDFG $G'$ such
that every execution sequence of  $G'$  is
an element of the set of execution sequences
of $A$.  Now, $G'$ and $G$ are shown to be isomorphic,
so that their sets of execution sequences are in 
a one-to-one correspondence with each other.  This
shows that the sets of execution sequences in $A$ and
$G$ are in a one-to-one correspondence.

@@

In each case, the VHDL generated by the C-to-RTL flow was synthesized
using Synposys Design Compiler, using the OSU Standard Cell
Library\footnote{\url{http://vcag.ecen.okstate.edu/projects/scells/}}
for the TSMC $0.18\mu$ technology. Memories were modeled using CACTI
SRAM
models\cite{CACTI}\footnote{\url{http://www.hpl.hp.com/research/cacti/}}.
The resulting net-list was mapped to layout using the Cadence SOC
Encounter tool. No human intervention was used in the entire process
from C to the final layout (other than specifying the operating
frequency to the automatic place-and-route tools). The total time
needed for the entire process (from C to layout) was of the order of a
few minutes for A5/1 to a few hours for the LPK case. A simulation
based toggle file was then used with the extracted net-list to
estimate power dissipation in each of the generated circuits. The
reciprocal of the energy used for completing a task is equivalent to
the throughput achieved for each watt of power supplied, commonly
termed as {\em performance per watt}

@@

As a reference to judge the energy efficiency of the AHIR generated
circuits, we consider a popular low power processor (henceforth
denoted by $P$), namely the Intel Atom N270
processor\footnote{\url{http://www.intel.com/products/processor/atom/index.htm}},
which is built in Intel's $45nm$ technology. This processor is readily
available, combines high performance with low power dissipation, has
full floating point support, and is a part of many low power computing
devices. The sample set of C programs was run on the processor $P$
(exactly the same program used in the hardware generation was run on
the processor). In each case, the number of cycles needed to finish
one task (as described above) specific to that program was measured.
The data sets for the programs were small enough to fit into the
processor cache, and the cycle count was averaged across multiple
runs. These cycle counts were then used to find the delay and energy
consumption for the processor. All values for area, frequency and
power dissipation for the processor were obtained from the
corresponding data-sheet.

@@

Now, suppose
that  we scale the 
energy numbers assuming migration of the AHIR circuits
from $180nm$ to the $45nm$ technology node, using 
the following  scaling rules (assume that $S = 45/180 = 0.25$):
\begin{itemize}
\item The operating voltage is scaled from $1.8V$ to $0.9V$.
\item The internal capacitances are assumed to be scaled by $S$.
\item The operating frequency of the generated circuit is scaled
by $1/S$.
\item Half of the total power dissipation of the $45nm$ circuits
is due to leakage\cite{ref:Frank}, and the other half is due to switching power. 
Note that this is a conservative estimate, and the actual leakage is
likely to be lower at the power levels corresponding to the
AHIR circuits.
%
% 
The leakage power is a negligible fraction of the
total power dissipation in the $180nm$ technology.
\item The area is scaled by $S^2$.
\end{itemize}
Thus, the per task energy consumption due to switching
activity alone will scale by $1/16$, and since
we are assuming that leakage power dissipation is half the total,
the total per task energy consumption of the circuit will 
scale by $1/8$.
The area of the circuits will scale by $1/16$.  
The power dissipation will scale by $1/2$ (this would have been
$1/4$ without leakage).
This scaling, though approximate, can be justified since the 
area of the resulting circuits (in each case) is small, so 
that interconnect effects are not likely to be pronounced.

@@

In Table \ref{table:Ratios}, we tabulate the ratios of
the performance parameters area, frequency, delay, power
and energy for the processor $P$ relative to the scaled values
for the AHIR circuits assuming the scaling rules shown above.
From the table, we see that across all applications, the energy consumed
by AHIR circuits is considerably lower than the
processor (the energy per task for the processor is
between $4.7X$ and $116.6X$ that of the corresponding AHIR circuit).  
The improvement is highest in the case of
bit-level manipulation applications such as A5/1, and
lowest in the case of random ``pointer-chasing'' code.
The delays in completing the tasks are lower 
for the processor $P$ in almost all the cases.  This
is mainly due to the fact that currently, the operators used in 
the AHIR generated RTL are not pipelined or optimized in any
manner during the logic synthesis and physical
design. This limits the operating clock frequency of the circuits
synthesized from the AHIR generated RTL.

@@
